---
title: "Building and solving a model"
subtitle: "A first script: a model file, a data frame, a solution"
format:
  html:
    toc: true
    toc-depth: 2
    embed-resources: true
execute:
  eval: false
---

A thortwo script always has the same three steps:

1. **Build** the model from its `.txt` file: `thor_model()`.
2. **Prepare** a data frame holding every variable of the model.
3. **Solve** it over a range of periods: `thor_solve()`.

The example below is complete: copy it into a script and run it.

## The complete script

The model, saved as `mini.txt` (see [Writing model equations](equations.qmd)
for the syntax):

```
endogenous :
ch, i, y

exogenous :
g

coefficients :
mpc, inv_share

equations :
y = ch + i + g
consumption : ch = mpc * lag(y, 1)
i = inv_share * y
```

The script:

```{r}
library(thortwo)

## 1. Build the model
model <- thor_model("mini", "mini.txt")

## 2. The data: one row per year, one column per variable
years <- 2015:2030
data <- data.frame(
  year      = years,
  g         = 20 * 1.02^(years - 2015),   # exogenous
  mpc       = 0.6,                        # coefficients
  inv_share = 0.25,
  y         = NA, ch = NA, i = NA         # endogenous: to be solved
)
data[data$year == 2015, c("y", "ch", "i")] <- c(100, 55, 25)   # history

## 3. Solve from 2016 to 2030
result <- thor_solve(model, from = 2016, to = 2030,
                     data = data, index_time = "year")

head(result, 3)
#>   year      g mpc inv_share       y    ch      i
#> 1 2015 20.000 0.6      0.25 100.000 55.00 25.000
#> 2 2016 20.400 0.6      0.25 107.200 60.00 26.800
#> 3 2017 20.808 0.6      0.25 113.504 64.32 28.376
```

`result` is your data frame with the endogenous variables filled in from 2016
to 2030. Everything else is left exactly as it was.

## The data frame

The data frame has **one row per period** and **one column per variable**,
plus one column for time.

| | |
|---|---|
| **Time column** | Any name, passed to `thor_solve()` as `index_time`. Years (`2019`), dates (`as.Date("2019-01-01")`) or labels (`"2019Q1"`) all work. No duplicates, no missing values. |
| **Row order** | **Sorted from oldest to newest, with no gaps.** A lag reads the previous *row*, so rows out of order give wrong results without any error. `time_model_checks(data, "year")` checks this. |
| **Column names** | Every variable of the model, in **lower case**, as in the model file. Extra columns are fine and are left untouched. |
| **Values** | Numbers. |

Which cells must be filled:

| variable | before the first solved period | over the solved periods |
|---|---|---|
| exogenous and coefficients | yes | **yes, every period** |
| endogenous | **yes** | no: they are what is solved |

- **History before the first solved period.** You need as many complete
  periods as the deepest lag in the model. `mini` uses `lag(y, 1)`, so one
  year is enough. thortwo prints how far back the model reaches when it is
  built, and refuses to start too early, naming the earliest possible period.
- **Endogenous values over the solved periods** are not needed. The solver
  starts each period from the previous period's solution, so it ignores
  whatever is there, `NA` included.
- **Coefficients** are ordinary columns, normally constant. If they come as a
  separate table of names and values, `add_coeffs()` adds them to the data:

  ```{r}
  coeffs <- data.frame(name = c("mpc", "inv_share"), value = c(0.6, 0.25))
  data <- add_coeffs(coeffs, data, pos.coeff.name = 1, pos.coeff.value = 2)
  ```

- `data_model_checks(model, data)` checks that every model variable is
  present.

## Building: `thor_model()`

```{r}
model <- thor_model("mini", "mini.txt")
```

That is all most scripts need. All the options:

| argument | default | what it does |
|---|---|---|
| `name` | (required) | A name for the model. It is also the name of the folder where its C++ code is kept, so use a different name for each model. |
| `source` | | Path to the `.txt` model file. |
| `backend` | `"sparse"` | How the model is solved: `"sparse"` (compiled), `"sparse-r"` (pure R, no compiler needed) or `"dense-cpp"`. See [Choosing a backend](backends.qmd). `"sparse"` is the right choice for normal work; `"sparse-r"` for a quick one-off run or a computer without a C++ compiler. |
| `decompose` | `TRUE` | Split the model into prologue, heart and epilogue blocks, which makes solving faster. `FALSE` solves it as one block, mostly useful for debugging. |
| `sequential` | `FALSE` | Advanced. `TRUE` solves the prologue and the epilogue one equation at a time instead of by Newton. See [below](#advanced-sequential-prologue-and-epilogue). |
| `compile` | `TRUE` | Compile straight away, so the model is ready to solve. With `FALSE`, compilation happens on the first `thor_solve()`. |
| `cache` | `NULL` | Where built and compiled models are kept between sessions. `NULL` uses the default cache; `FALSE` turns it off. |
| `recompile` | `FALSE` | `TRUE` builds and compiles from scratch even if this model is in the cache. |
| `workdir` | `NULL` | Where the generated code is written. Leave it alone: the default is what lets the cache recognise a model it has already compiled. |
| `verbose` | `TRUE` | Print progress. |
| `timings` | `TRUE` | Print how long the build and the compilation took, even with `verbose = FALSE`. Also kept in `m@meta$timings`. |

Instead of a file, the model can be given as vectors, through `endogenous`,
`exogenous`, `coefficients` and `equations`.

### Built models are cached

Building a large model takes time. So every model you build is stored, and
when you call `thor_model()` again on a model that has not changed, in the
same R session or weeks later, it is loaded instead of being rebuilt:

```
This version of the model has already been compiled before; using the existing cache.
To recompile, use recompile = TRUE and rerun the build.
```

On ThreeME 13x13 that takes half a second instead of about a minute.

- **What counts as "the same model":** the same name, variables, equations,
  `backend` and `decompose`, built with the same version of thortwo and of R.
  Change an equation, even one number in it, and the model is built again.
  Spacing, upper or lower case, and the name or date of the file don't
  matter.
- **To force a fresh build:** `thor_model(..., recompile = TRUE)`.
- **To switch the lookup off:** `cache = FALSE` for one call, or
  `options(thortwo.model.cache = FALSE)` for the session.
- **To empty the cache:** `clear_model_cache()`. It lives in a per-user
  folder, shown by `thor_cache_dir()`, and takes about 1 MB per model plus the
  compiled code.

### Advanced: sequential prologue and epilogue

```{r}
model <- thor_model("mymodel", "mymodel.txt", sequential = TRUE)
```

The decomposition splits a model into three blocks. The prologue and the
epilogue are *recursive*: taken in the right order, each of their equations
has only one variable that is not yet known. By default thortwo still solves
them like the heart, as a system, with Newton's method. With
`sequential = TRUE` it solves them one equation at a time, in that order:

- an equation written `x = f(...)`, where `x` is the variable it determines
  and doesn't appear on the right, is simply **evaluated**;
- any other equation (`log(y) = ...`, `pk * f_k = ...`) is solved for its one
  unknown with a **one-variable Newton** iteration.

The heart is still solved by Newton. `thor_solve()` is called exactly as
before, and printing the model shows what was done:

```
  prologue    792 equations   sequential: 750 direct, 42 by one-variable Newton
  heart      4663 equations   jacobian 16513 non-zero (0.08% dense)
  epilogue   1336 equations   sequential: 1087 direct, 249 by one-variable Newton
```

**What it changes.** The solution is the same as Newton's, to within the
solver's tolerance. Measured on the `"sparse"` backend:

| | normal | sequential |
|---|---|---|
| ThreeME 4x4: build in R / compile / solve | 6.4 s / 7.5 s / 0.15 s | 7.2 s / 8.3 s / 0.14 s |
| ThreeME 13x13: build in R / compile / solve | 29 s / 18 s / 3.8 s | 26 s / 17 s / 4.1 s |
| Opale: solve | 0.011 s | 0.003 s |

So on these models the effect on time is **small**, in either direction. No
jacobian is derived or compiled for the two blocks, but their equations are
written out a second time in assignment form, and the two roughly cancel. The
pure-R backend gains a little more (13x13 solve: 19 s to 16 s).

Where it does help:

- **Clearer errors.** When a variable can't be determined, the message names
  the equation and the variable instead of reporting that a whole block did
  not converge:

  ```
  Sequential solve failed on block 'prologue' at 2021 (row 3): equation 'eq_2528' could
  not be solved for 'pk_sgzx'. Its derivative with respect to that variable is
  zero or not finite, or it did not converge in 100 iterations.
  ```

- **Probably, models whose prologue or epilogue is large compared with the
  heart.** These blocks are then no longer solved as systems. This has not
  been measured on such a model.

It is off by default, and with it off nothing changes. It has no effect with
`decompose = FALSE`.

## Solving: `thor_solve()`

```{r}
result <- thor_solve(model, from = 2016, to = 2030,
                     data = data, index_time = "year")
```

| argument | default | what it does |
|---|---|---|
| `model` | (required) | The model from `thor_model()`. |
| `from`, `to` | (required) | First and last period to solve, as they appear in the time column. |
| `data` | (required) | The data frame. |
| `index_time` | `"date"` | Name of the time column. |
| `rtol` | `1e-10` | How precisely each variable is solved, relative to its size. |
| `atol` | `1e-8` | The same, for variables at or near zero. |
| `max_iter` | `100` | Maximum Newton iterations per block and period before giving up. |
| `damping` | `TRUE` | Shorten a Newton step that would make things worse. Leave it on. |
| `reuse_jacobian` | on, except on `"sparse-r"` | Reuse the factorised jacobian within a period, which makes large models solve about three times faster. `FALSE` gives plain Newton. See [below](#reusing-the-jacobian). |
| `verbose` | `TRUE` | Print progress. |
| `timings` | `TRUE` | With `verbose = FALSE`, print how long the solve took. With `verbose = TRUE` the summary line already says. |
| `diagnostics` | `FALSE` | Attach per-period details to the result (below). |

The solver works period by period. Within a period it solves the prologue, then
the heart, then the epilogue, each with Newton's method, starting from the
previous period's values.

With `diagnostics = TRUE`, the result carries:

```{r}
attr(result, "iterations")    # Newton iterations, per period
attr(result, "residuals")     # largest equation error, per period
attr(result, "convergence")   # worst scaled step, per period (below 1 = converged)
attr(result, "elapsed")       # solve time in seconds
```

### Reusing the jacobian

At every Newton iteration the solver has to prepare the model's table of
derivatives (the jacobian) before it can compute a step. This preparation,
the *factorisation*, is by far the most expensive part of an iteration on a
large model: 1.4 seconds on ThreeME 29x33, against a few milliseconds for
everything else.

So by default the solver factorises once per period and reuses the result for
that period's other iterations. A step computed with a slightly out-of-date
factorisation is a little less accurate, so more steps are needed, but each
one costs almost nothing. If such a step fails to reduce the error, it is
discarded and redone the ordinary way, so the worst case is plain Newton.

| `"sparse"` backend | plain Newton | reusing (default) |
|---|---|---|
| ThreeME 13x13, 38 years | 3.8 s, 456 iterations | **1.2 s**, 996 iterations |
| ThreeME 29x33, 38 years | 208 s, 456 iterations | **74 s**, 1,009 iterations |
| Opale, 40 quarters | 0.011 s | 0.009 s |

The solution is the same: on these models it differs from plain Newton's by
less than `1e-10`, and the equations are satisfied just as closely.

- **The number of iterations reported is about twice what plain Newton
  takes.** That is expected: more steps, each much cheaper.
- **It is off by default on the `"sparse-r"` backend.** There, evaluating the
  equations in R is the main cost, and more iterations means more of it.
- To get plain Newton, for instance to compare iteration counts with an
  earlier run:

  ```{r}
  result <- thor_solve(model, from, to, data = data, index_time = "year",
                       reuse_jacobian = FALSE)
  ```

## Checking the result

A converged solve means Newton stopped moving. To check that the equations
actually hold:

```{r}
model_residuals(model, result, periods = 2016:2030, index_time = "year")
#>          prologue heart
#> 2016 2.220446e-15     0
#> 2017 2.380318e-15     0
```

This gives the largest equation error per block and period. Values like
`2e-15` are rounding; anything around `1e-8` or below is fine. Add `by = "equation"` to see every equation.

`calibration_check()` lists the equations that a data frame doesn't satisfy in
one period. It is most useful **before** solving, on a calibrated baseline,
to find equations the calibration misses; that is the first thing to run when
a solve fails. The period needs the same history as a solve does.

```{r}
calibration_check(model, result, period = 2017, index_time = "year")
#> [1] equation part     formula  residual
#> <0 rows> (or 0-length row.names)
```

No rows means every equation holds to within `1e-6`.

## Saving and reloading a model

```{r}
thor_save(model, "mini.rds")

## later, in another session or on another computer
model <- thor_load("mini.rds")
```

The saved file contains everything: the equations and the generated code. It
doesn't need the original `.txt` file, and on the same computer it reloads from
the cache without compiling again.

## When something goes wrong

| symptom | where to look |
|---|---|
| Building stops with an error about an equation | [Writing model equations](equations.qmd), section 7 |
| `Missing variables in the database: …` | a column is missing or not in lower case |
| `The first period to solve is … but this model reaches … back` | start later, or add history to the data |
| `… missing over the solved period: …` | an exogenous variable or coefficient has an `NA` |
| `Newton did not converge on block …` | run `calibration_check()` on that period to find the equations your data misses |
| `Sequential solve failed … equation '…' could not be solved for '…'` | with `sequential = TRUE`: that equation doesn't determine that variable at that period, typically because something multiplying it is zero in the data |
| Compiler errors mentioning `math.h` or `cstdint` | your computer's C++ setup: run `thor_check_toolchain()` |
