---
title: "Solving ThreeME with thortwo"
subtitle: "Building a model from a text file, solving it, and reloading it later"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    df-print: kable
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", dev = "png", dpi = 160)
```

```{r libraries}
#| message: false
library(thortwo)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
```

::: {.callout-note collapse="true"}
## A note on dependencies

`thortwo` itself depends on no part of the tidyverse — deliberately, since the
solver has no use for it. `dplyr`, `tidyr`, `purrr` and `ggplot2` are used *in this
document* to handle and show results, which is how you would work with them in
practice.
:::

## What we are doing

Two workflows, end to end:

1. **Build and solve.** Read a ThreeME model from its `.txt` file, build it,
   load the data, solve it over a projection horizon, and check the answer.
2. **Reload.** Save the built model, and pick it up again in a later session
   without paying for the build a second time.

ThreeME at the 4x4 classification is 1729 equations. Everything below runs in
well under a minute on a laptop.

```{r files}
model_file <- system.file("ThreeME", "threeme_4x4_thor.txt", package = "thortwo")
data_file  <- system.file("ThreeME", "data3me_4x4.rds",      package = "thortwo")

c(model = basename(model_file), data = basename(data_file))
```

## 1. The model file

A model file is four sections: the endogenous variables, the exogenous
variables, the coefficients, and the equations. Section headers are what
identify them, so blank lines and separator banners are free — you cannot
break the file by adding a line in the wrong place.

```{r peek-file}
writeLines(substr(head(readLines(model_file), 11), 1, 78))
```

You can read those sections without building anything, which is the quickest
way to check that a file is well formed:

```{r read-source}
parts <- read_model_source(model_file)

tibble(
  section = names(parts),
  n       = lengths(parts)
)
```

An equation is written in the obvious way, optionally with a name in front of
a colon. `lag(x, n)` and `delta(n, x)` are understood, along with the usual
mathematical functions:

```{r peek-equations}
writeLines(substr(parts$equations[1:4], 1, 78))
```

## 2. Building the model

`thor_model()` does the whole pipeline: check, decompose into blocks,
differentiate symbolically, generate a solver, compile it.

```{r build}
#| cache: false
t_build <- system.time(
  m <- thor_model("threeme4x4", model_file, backend = "sparse", verbose = FALSE,
                  recompile = TRUE)
)

m
```

`recompile = TRUE` is there only so that this page shows a real build every
time it is rendered. Without it, `thor_model()` recognises a model it has
already built and loads it from its cache instead, in about half a second,
with a message saying so (see [the cache](#the-cache) below).

```{r build-time}
cat("built in", round(t_build[["elapsed"]], 1), "s\n")
```

### What the decomposition bought

The model is split into three blocks that are solved in sequence rather than
all at once. Newton's cost is cubic in block size, so this matters a great
deal: only the *heart* is a genuinely simultaneous system.

```{r blocks}
tibble(
  block            = names(m@blocks),
  equations        = map_int(m@blocks,   ~ length(.x$equations)),
  jacobian_entries = map_int(m@jacobian, ~ length(.x$i))
) |>
  mutate(pct_dense = round(100 * jacobian_entries / equations^2, 3))
```

Around a third of one percent of the heart's jacobian is non-zero — 4160
entries in a 1107 x 1107 matrix. That is the whole argument for the `sparse`
backend: the dense representation would allocate and factorise all 1.2 million
cells on every Newton iteration to carry those 4160.

### Choosing a backend

`backend` is stored on the model, and `thor_solve()` dispatches on it, so
there is only ever one way to solve.

| backend | jacobian | needs a compiler | use it when |
|---|---|---|---|
| `"sparse"` | triplets, Eigen `SparseLU` | yes | the default: the fastest solve |
| `"sparse-r"` | triplets, `Matrix` sparse LU, pure R | no | no toolchain, or one quick run |
| `"dense-cpp"` | dense, Eigen `PartialPivLU` | yes | a cross-check of `"sparse"` |

All three produce the same answer — on Opale they agree to a median relative
difference of about 1e-15 — but on ThreeME 4x4 the sparse solver is roughly
30x faster than the dense one, and at 8x8 it is 70x.

## 3. The data

One row per period, one column per variable, plus a column to use as the time
index.

```{r data}
data_3me <- readRDS(data_file)

cat(nrow(data_3me), "periods x", ncol(data_3me), "columns\n")
cat("years", min(data_3me$year), "to", max(data_3me$year), "\n")

data_3me |> select(1:5) |> slice_head(n = 4)
```

Before solving, it is worth asking whether the database actually holds
everything the model needs:

```{r data-checks}
data_model_checks(m, data_3me, quiet = FALSE)
```

If your coefficients live outside the database — as they do for Opale, in a
separate two-column table — `add_coeffs()` broadcasts them on as constant
columns first. ThreeME carries its own, so nothing to do here.

## 4. Solving

`thor_solve()` runs Newton per block, per period, starting each period from
the previous one.

Two things to know about the period bounds. The first solved period cannot be
the first row of the data: the solver needs a complete previous observation to
initialise from and to supply the lagged terms. And every exogenous variable
and coefficient must be present over the whole span — if one is missing,
you get told which, rather than a convergence failure twenty periods later.

```{r solve}
res <- thor_solve(m, from = 2016, to = 2050,
                  data = data_3me, index_time = "year",
                  verbose = FALSE, diagnostics = TRUE)
```

`diagnostics = TRUE` attaches the per-period record:

```{r diagnostics}
tibble(
  year        = as.integer(names(attr(res, "iterations"))),
  iterations  = as.integer(attr(res, "iterations")),
  scaled_step = signif(attr(res, "convergence"), 3),
  residual    = signif(attr(res, "residuals"), 3)
) |>
  slice_head(n = 8)
```

```{r solve-summary}
cat("solved", length(attr(res, "iterations")), "periods in",
    round(attr(res, "elapsed"), 3), "s\n")
cat("total Newton iterations:", sum(attr(res, "iterations")), "\n")
cat("worst scaled step:", signif(max(attr(res, "convergence")), 3),
    "(converges at 1)\n")
```

The scaled step is a mixed criterion — a variable has converged when
`|dx| <= rtol*|x| + atol` — reported normalised, so anything at or below 1 has
converged. It is written that way because macro models mix variables of wildly
different magnitude, and a great many of them sit at exactly zero in a given
scenario.

### Checking the answer

Convergence is not the same as correctness. Newton will happily converge on
whatever equations it was given; only the residuals notice if those are not
the equations in your file. `model_residuals()` re-evaluates every equation at
the solution:

```{r residuals}
resid <- model_residuals(m, res, periods = 2016:2050, index_time = "year")

round(apply(resid, 2, max), 12)
```

```{r residuals-check}
stopifnot(max(resid) < 1e-6)
cat("max |residual| over all blocks and periods:", signif(max(resid), 3), "\n")
```

Make this check a habit. It is the one that catches code-generation bugs, and
it costs milliseconds.

### Looking at the result

`res` is the input data frame with the endogenous variables filled in over the
solved span; everything else is untouched.

```{r results}
vars <- c("gdp", "pgdp", "unr", "ems_co2")

res |>
  filter(year %in% seq(2016, 2050, by = 5)) |>
  select(year, all_of(vars))
```

## 5. Running a scenario

The usual reason to solve a model twice: change an exogenous input and compare.
A scenario is an ordinary edit to the data frame — there is no separate shock
machinery to learn. Here we raise the rate of social contributions by 10% from
2025 on.

```{r scenario}
shock_var <- "rinc_soc_tax"

## It has to be exogenous: solving would simply overwrite an endogenous one.
stopifnot(shock_var %in% m@vars$exo)

## Flat across the baseline, so the shock below is a clean level change.
## (`unique()` would report 31 values here: they differ only in the last bit.)
summary(data_3me[[shock_var]])
```

```{r scenario-solve}
data_shock <- data_3me
rows <- data_shock$year >= 2025
data_shock[rows, shock_var] <- data_shock[rows, shock_var] * 1.10

res_shock <- thor_solve(m, 2016, 2050, data_shock, "year", verbose = FALSE)
```

Both runs used the same model object, so the comparison is like for like —
only the input changed.

```{r scenario-compare}
reported <- c(gdp = "GDP", unr = "Unemployment rate", ems_co2 = "CO2 emissions")

compare <- bind_rows(
    baseline = res      |> select(year, all_of(names(reported))),
    scenario = res_shock |> select(year, all_of(names(reported))),
    .id = "run"
  ) |>
  pivot_longer(all_of(names(reported)), names_to = "variable", values_to = "value") |>
  pivot_wider(names_from = run, values_from = value) |>
  mutate(
    pct_diff = 100 * (scenario / baseline - 1),
    variable = factor(reported[variable], levels = unname(reported))
  ) |>
  ## Only the solved window is meaningful: outside it the two runs hold the
  ## same untouched data, so the difference is identically zero.
  filter(year >= 2016, year <= 2050)

compare |>
  filter(year %in% seq(2025, 2050, by = 5)) |>
  select(year, variable, pct_diff) |>
  pivot_wider(names_from = variable, values_from = pct_diff) |>
  mutate(across(-year, \(x) round(x, 3)))
```

Read as percentage deviations from baseline. The table is the accessible view
of the chart below — every plotted value is here.

All three series are in the same unit — percent deviation from baseline — so
they belong on one axis. (Two measures of *different* scale would need two
charts, or indexing to a common base; never a second y-axis.)

One caveat on reading it: the unemployment rate is itself a rate, so its
deviation is relative — it rises from about 10.4% to 11.3% at the peak, which is
+9% relative but under one percentage point.

```{r scenario-plot}
#| fig-width: 8
#| fig-height: 4.5
#| fig-alt: "Line chart of percentage deviation from baseline for GDP, the unemployment rate and CO2 emissions, 2016 to 2050, under a 10 percent higher rate of social contributions from 2025. All three move away from zero after the 2025 shock and the lines are directly labelled at the right."
## Categorical palette slots 1-3, validated for colour-vision deficiency
## separation against the light chart surface.
series_colours <- c("GDP"               = "#2a78d6",
                    "Unemployment rate" = "#eb6834",
                    "CO2 emissions"     = "#1baf7a")

ends <- compare |> filter(year == max(year))

ggplot(compare, aes(year, pct_diff, colour = variable)) +
  geom_hline(yintercept = 0, linewidth = 0.4, colour = "#b5b3ad") +
  geom_vline(xintercept = 2025, linewidth = 0.4, linetype = "dotted",
             colour = "#b5b3ad") +
  geom_line(linewidth = 0.8) +
  ## Direct labels as well as the legend, so identity is never colour alone.
  ## The label text wears ink rather than the series colour: the line it sits
  ## against carries the identity, and coloured text would be the harder read.
  geom_text(data = ends, aes(label = variable), hjust = 0, nudge_x = 0.6,
            size = 3.2, colour = "#0b0b0b", show.legend = FALSE) +
  annotate("text", x = 2025, y = Inf, label = "shock", vjust = 1.4, hjust = -0.15,
           size = 3, colour = "#52514e") +
  scale_colour_manual(values = series_colours, name = NULL) +
  scale_x_continuous(limits = c(2016, 2064), breaks = seq(2020, 2050, 10)) +
  scale_y_continuous(labels = scales::label_number(suffix = "%")) +
  labs(title = "A 10% higher rate of social contributions from 2025",
       subtitle = "Deviation from baseline, ThreeME 4x4",
       x = NULL, y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    plot.background   = element_rect(fill = "#fcfcfb", colour = NA),
    panel.background  = element_rect(fill = "#fcfcfb", colour = NA),
    panel.grid.minor  = element_blank(),
    panel.grid.major  = element_line(colour = "#e8e7e1", linewidth = 0.3),
    plot.title        = element_text(size = 11.5, colour = "#0b0b0b"),
    plot.subtitle     = element_text(size = 9.5,  colour = "#52514e"),
    axis.text         = element_text(colour = "#52514e"),
    legend.position   = "top",
    legend.justification = "left",
    legend.text       = element_text(colour = "#52514e")
  )
```

## 6. Saving and reloading a model

Building costs seconds to minutes. Solving costs milliseconds. So you build
once and reload after that.

A saved model is **self-contained**: the `.rds` carries the generated solver
source inside it, not a path to it. You can copy the file to another machine,
commit it, or come back to it after a reboot has cleared `/tmp`, and it will
still work.

```{r save}
store <- file.path(tempdir(), "threeme4x4.rds")   # somewhere durable, in practice
thor_save(m, store)

cat("saved", round(file.size(store) / 1024^2, 1), "MB\n")
```

### In a later session

```{r load}
t_load <- system.time(
  m2 <- thor_load(store)
)

m2
cat("loaded and ready in", round(t_load[["elapsed"]], 2), "s\n")
```

That is the whole reload workflow — there is no separate "recompile" step to
remember. `thor_load()` writes the stored source out and compiles it, which
hits the compile cache whenever the code has not changed, so in practice it
costs a fraction of a second. A model built here and a model loaded from disk
are the same object:

```{r load-verify}
res2 <- thor_solve(m2, 2016, 2050, data_3me, "year", verbose = FALSE)

identical(res[, m@vars$endo], res2[, m2@vars$endo])
```

### The cache

The cache is what makes the above cheap, and it is on by default. It keeps
two things: every **built model**, under a hash of its name, variables,
equations and options, and its **compiled code**. So calling `thor_model()`
again on a model whose equations have not changed, in this session or a later
one, neither rebuilds nor recompiles it: it is loaded, and a message says so.
`thor_model(..., recompile = TRUE)` forces a fresh build.

```{r cache}
thor_cache_dir()
```

```r
options(thortwo.cache.dir = "~/my-cache")   # move it
options(thortwo.cache.dir = FALSE)          # turn it off
clear_model_cache()                         # empty it
thor_model(..., cache = FALSE)              # or just for this build
options(thortwo.model.cache = FALSE)        # compile cache only: always rebuild the model
```

If you edit the model file, the generated code changes, the hash changes, and
the next build recompiles by itself. You never have to remember to invalidate
anything.

## 7. Where to go wrong

A few failure modes worth recognising:

**"The first period cannot be the first observation."** The solver initialises
from the previous row. Start at the second row or later.

**"Exogenous variables or coefficients are missing over the solved period."**
Exactly what it says, with the names listed. Extend your exogenous paths to
cover the whole horizon.

**"Newton did not converge on block 'heart' at 2021 (row N)."** The message reports
the scaled step actually reached, the tolerance, and the worst residual. In
order of likelihood: bad data at that period, an over-tight `rtol`, or a model
that genuinely has no solution there. Try `max_iter = 200`, or look at the
data for that year first.

**Converged, but the residuals are large.** That is a code-generation problem,
not a numerical one. Report it — and it is why section 4 checks.

## Summary

```r
library(thortwo)

# once
m <- thor_model("mymodel", "mymodel.txt")
thor_save(m, "mymodel.rds")

# every time after that
m   <- thor_load("mymodel.rds")
res <- thor_solve(m, from = 2016, to = 2050, data = mydata, index_time = "year")
stopifnot(max(model_residuals(m, res, 2016:2050, "year")) < 1e-6)
```

```{r session}
#| echo: false
cat("thortwo", as.character(packageVersion("thortwo")),
    "on", R.version.string, "\n")
```
